---
title: "Taxonomy_phage_bacteria"
output: html_document
date: "2025-12-22"
---
```{r}
## FOR PHAGES:
suppressPackageStartupMessages({
  library(here)
  library(dplyr); library(readr); library(tidyr); library(stringr)
  library(ggplot2); library(scales)
})

# ===================== PATHS =====================
# Base output dir inside the repository
out_dir <- here("results", "taxonomy")
fig_dir <- file.path(out_dir, "figures")
tab_dir <- file.path(out_dir, "tables")

dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)

# Input data inside the repository
phage_results <- here("data", "results_with_metadata.csv")

# ---- Load data ----
df <- readr::read_csv(phage_results, show_col_types = FALSE)

# ===================== PALETTE =====================
user_palette_pastel <- c(
  "Streptococcus" = "pink",
  "Bacteroides"   = "mediumaquamarine",
  "Escherichia"   = "gold",
  "Faecalibacterium" = "#C499CC",
  "Bifidobacterium"  = "#6FA3F7",
  "Klebsiella"    = "#7B2CBF",
  "Blautia"       = "#90DDF0",
  "Gemmiger"      = "#FFAC81",
  "Ruminococcus"  = "#FA6459",
  "Pseudomonas"   = "#EEE2DF"
)
user_palette <- user_palette_pastel

# ===================== HELPERS =====================
extract_clean_genus <- function(x){
  g <- ifelse(!is.na(x) & str_detect(x, "g__"),
              str_replace(x, ".*g__", ""),
              NA_character_)
  g <- sub("_[A-Z]$", "", g)
  dplyr::recode(g,
    "Kleibsella" = "Klebsiella",
    "Gemminger"  = "Gemmiger",
    .default = g
  )
}

base_prepare <- function(df){
  df %>%
    mutate(
      Group = dplyr::case_when(
        mother_infant == "infant" & timepoint_months == 0.25 ~ "1 week",
        mother_infant == "infant" & timepoint_months == 1    ~ "1 month",
        mother_infant == "infant" & timepoint_months == 6    ~ "6 months",
        mother_infant == "infant" & timepoint_months == 12   ~ "12 months",
        mother_infant == "mother" & bodysite == "stool"      ~ "mothers",
        bodysite == "breastmilk"                              ~ "milk",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(Group)) %>%
    filter(CPM > 0) %>%
    mutate(Genus = extract_clean_genus(Host.genus)) %>%
    filter(!is.na(Genus))
}

ensure_palette <- function(levels_vec, pal){
  miss <- setdiff(levels_vec, names(pal))
  if(length(miss) > 0){
    pal <- c(pal, setNames(scales::hue_pal()(length(miss)), miss))
  }
  pal
}

# ===================== PLOTS =====================
plot_top10 <- function(df, include_other = FALSE,
                       title = "Average proportional distribution of top-10 predicted host genera (CPM)"){

  dat <- base_prepare(df)

  cpm_by_sample_genus <- dat %>%
    group_by(Sample, Group, Genus) %>%
    summarise(total_CPM = sum(CPM), .groups = "drop")

  top10 <- cpm_by_sample_genus %>%
    group_by(Genus) %>%
    summarise(global_CPM = sum(total_CPM), .groups = "drop") %>%
    arrange(desc(global_CPM)) %>%
    slice_head(n = 10) %>%
    pull(Genus)

  df_top <- cpm_by_sample_genus %>%
    mutate(Genus = ifelse(Genus %in% top10, Genus, "Other"))

  prop <- df_top %>%
    group_by(Sample) %>%
    mutate(Proportion = total_CPM / sum(total_CPM)) %>%
    ungroup()

  group_mean <- prop %>%
    group_by(Group, Genus) %>%
    summarise(Mean_Proportion = mean(Proportion), .groups = "drop")

  group_mean_norm <- group_mean %>%
    group_by(Group) %>%
    mutate(Mean_Proportion = Mean_Proportion / sum(Mean_Proportion)) %>%
    ungroup()

  if(!include_other){
    group_mean_norm <- group_mean_norm %>%
      filter(Genus != "Other") %>%
      group_by(Group) %>%
      mutate(Mean_Proportion = Mean_Proportion / sum(Mean_Proportion)) %>%
      ungroup()
  }

  group_levels <- c("1 week","1 month","6 months","12 months","mothers","milk")
  group_mean_norm$Group <- factor(group_mean_norm$Group, levels = group_levels)

  genus_order <- group_mean_norm %>%
    group_by(Genus) %>%
    summarise(tot = sum(Mean_Proportion), .groups = "drop") %>%
    arrange(desc(tot)) %>%
    pull(Genus)

  if("Other" %in% genus_order){
    genus_order <- c("Other", setdiff(genus_order, "Other"))
  }
  group_mean_norm$Genus <- factor(group_mean_norm$Genus, levels = genus_order)

  pal <- ensure_palette(levels(group_mean_norm$Genus), user_palette)
  if("Other" %in% names(pal)) pal["Other"] <- "grey70"

  ggplot(group_mean_norm, aes(Group, Mean_Proportion, fill = Genus)) +
    geom_col(color = "black", linewidth = 0.2) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_manual(values = pal, drop = FALSE) +
    labs(x = NULL, y = "Relative abundance (%)", fill = "Predicted host genus", title = title) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom",
          legend.title = element_text(face = "bold"))
}

plot_all <- function(df, title = "Average proportional distribution of predicted host genera (CPM) — ALL"){
  dat <- base_prepare(df)

  cpm_by_sample_genus <- dat %>%
    group_by(Sample, Group, Genus) %>%
    summarise(total_CPM = sum(CPM), .groups = "drop")

  prop <- cpm_by_sample_genus %>%
    group_by(Sample) %>%
    mutate(Proportion = total_CPM / sum(total_CPM)) %>%
    ungroup()

  group_mean_norm <- prop %>%
    group_by(Group, Genus) %>%
    summarise(Mean_Proportion = mean(Proportion), .groups = "drop") %>%
    group_by(Group) %>%
    mutate(Mean_Proportion = Mean_Proportion / sum(Mean_Proportion)) %>%
    ungroup()

  group_levels <- c("1 week","1 month","6 months","12 months","mothers","milk")
  group_mean_norm$Group <- factor(group_mean_norm$Group, levels = group_levels)

  genus_order <- group_mean_norm %>%
    group_by(Genus) %>%
    summarise(tot = sum(Mean_Proportion), .groups = "drop") %>%
    arrange(desc(tot)) %>%
    pull(Genus)
  group_mean_norm$Genus <- factor(group_mean_norm$Genus, levels = genus_order)

  pal <- ensure_palette(levels(group_mean_norm$Genus), user_palette)

  ggplot(group_mean_norm, aes(Group, Mean_Proportion, fill = Genus)) +
    geom_col(color = "black", linewidth = 0.2) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_manual(values = pal, drop = FALSE) +
    labs(x = NULL, y = "Relative abundance (%)", fill = "Predicted host genus", title = title) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          axis.text.x = element_text(angle = 30, hjust = 1),
          legend.position = "bottom",
          legend.title = element_text(face = "bold"))
}

# ===================== RUN =====================
p_top10_noOther   <- plot_top10(df, include_other = FALSE)
p_top10_withOther <- plot_top10(df, include_other = TRUE)
p_all             <- plot_all(df)

# Optional: show one
print(p_top10_noOther)

# ===================== SAVE (INSIDE REPO) =====================
ggsave(filename = file.path(fig_dir, "top10_NO_OTHER.png"),
       plot = p_top10_noOther, width = 11, height = 10, dpi = 300)

ggsave(filename = file.path(fig_dir, "top10_WITH_OTHER.png"),
       plot = p_top10_withOther, width = 11, height = 10, dpi = 300)

ggsave(filename = file.path(fig_dir, "ALL_genera.png"),
       plot = p_all, width = 11, height = 10, dpi = 300)

# Optional: save the plotted data too (reproducibility)
# write_csv(..., file.path(tab_dir, "top10_NO_OTHER_plotdata.csv"))
```



```{r}
## FOR BACTERIA (MMC3) — same style as your PHAGE plot
suppressPackageStartupMessages({
  library(here)
  library(dplyr); library(tidyr); library(stringr)
  library(readxl)
  library(ggplot2); library(scales)
  library(readr)
})

# ===================== PATHS =====================
out_dir <- here("results", "taxonomy_bacteria_mmc3_like_phage")
fig_dir <- file.path(out_dir, "figures")
tab_dir <- file.path(out_dir, "tables")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)

# Inputs
bact_xlsx <- here("data", "mmc3.xlsx")
meta_csv  <- here("data", "results_with_metadata.csv")  # to build Group + bodysite

# ===================== PALETTE =====================
user_palette_pastel <- c(
  "Streptococcus"     = "pink",
  "Bacteroides"       = "mediumaquamarine",
  "Escherichia"       = "gold",
  "Faecalibacterium"  = "#C499CC",
  "Bifidobacterium"   = "#6FA3F7",
  "Klebsiella"        = "#7B2CBF",
  "Blautia"           = "#90DDF0",
  "Gemmiger"          = "#FFAC81",
  "Ruminococcus"      = "#FA6459",
  "Pseudomonas"       = "#EEE2DF"
)
user_palette <- user_palette_pastel

ensure_palette <- function(levels_vec, pal){
  miss <- setdiff(levels_vec, names(pal))
  if(length(miss) > 0){
    pal <- c(pal, setNames(scales::hue_pal()(length(miss)), miss))
  }
  pal
}

# ===================== HELPERS =====================
extract_genus_tax <- function(x){
  ifelse(!is.na(x) & str_detect(x, "g__"),
         str_replace(x, ".*g__", ""),
         NA_character_)
}

# Collapse MMC3 variants like Blautia_A, Ruminococcus_B, etc -> Blautia, Ruminococcus
collapse_to_palette <- function(genus_vec, palette_keys){
  out <- as.character(genus_vec)
  out <- str_trim(out)
  out <- str_replace_all(out, "\\s+", " ")

  # Remove common suffix patterns
  # Blautia_A, Blautia_B, Blautia_*, Ruminococcus_B -> Blautia, Ruminococcus
  out <- str_replace(out, "_.*$", "")

  # Force mapping if starts with palette key
  for(k in palette_keys){
    out <- ifelse(str_detect(out, paste0("^", k, "($|[_\\-].*)")), k, out)
  }

  # Common typos
  out <- dplyr::recode(out,
    "Kleibsella" = "Klebsiella",
    "Gemminger"  = "Gemmiger",
    .default = out
  )

  out
}

# OPTIONAL: if you want Phocaeicola to be counted as Bacteroides for visuals
# (MetaPhlAn often uses Phocaeicola as a renamed Bacteroides genus)
MAP_PHOC_TO_BACT <- FALSE

# ===================== LOAD METADATA (Group + bodysite) =====================
meta_src <- readr::read_csv(meta_csv, show_col_types = FALSE)

meta <- meta_src %>%
  select(Sample, mother_infant, bodysite, timepoint_months) %>%
  distinct() %>%
  mutate(
    Group = dplyr::case_when(
      mother_infant == "infant" & timepoint_months == 0.25 ~ "1 week",
      mother_infant == "infant" & timepoint_months == 1    ~ "1 month",
      mother_infant == "infant" & timepoint_months == 6    ~ "6 months",
      mother_infant == "infant" & timepoint_months == 12   ~ "12 months",
      mother_infant == "mother" & bodysite == "stool"      ~ "mothers",
      bodysite == "breastmilk"                             ~ "milk",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Group))

# ===================== LOAD MMC3 =====================
bact_raw <- readxl::read_excel(bact_xlsx)
stopifnot("clade_name" %in% names(bact_raw))

# MMC3 often uses comma as decimal separator
bact_num <- bact_raw %>%
  mutate(across(-clade_name, ~ as.numeric(gsub(",", ".", as.character(.)))))

bact_long <- bact_num %>%
  pivot_longer(-clade_name, names_to = "Sample", values_to = "percent") %>%
  filter(!is.na(percent)) %>%
  mutate(prop = percent / 100)

palette_keys <- names(user_palette)

bact_genus <- bact_long %>%
  mutate(Genus_raw = extract_genus_tax(clade_name)) %>%
  filter(!is.na(Genus_raw)) %>%
  mutate(
    Genus = collapse_to_palette(Genus_raw, palette_keys),
    Genus = ifelse(MAP_PHOC_TO_BACT & Genus == "Phocaeicola", "Bacteroides", Genus)
  ) %>%
  group_by(Sample, Genus) %>%
  summarise(Rel_abund_bact = sum(prop, na.rm = TRUE), .groups = "drop")

# ===================== PHAGE-STYLE BASE PREP =====================
base_prepare_bacteria <- function(df_bact, meta_df, stool_only = TRUE){
  out <- df_bact %>%
    inner_join(meta_df, by = "Sample") %>%
    filter(!is.na(Group)) %>%
    filter(!is.na(Genus)) %>%
    filter(Rel_abund_bact > 0)

  if(stool_only){
    out <- out %>% filter(bodysite == "stool")
  }
  out
}

# ===================== PLOTS (same logic as phage) =====================
plot_top10_bacteria <- function(df_bact, meta_df, stool_only = TRUE,
                                include_other = FALSE,
                                title = "Average proportional distribution of top-10 bacterial genera (MMC3)"){
  dat <- base_prepare_bacteria(df_bact, meta_df, stool_only = stool_only)

  # Abundance by Sample x Genus (already proportional, but we keep "total per genus per sample")
  abund_by_sample_genus <- dat %>%
    group_by(Sample, Group, Genus) %>%
    summarise(total_abund = sum(Rel_abund_bact), .groups = "drop")

  # Top-10 global
  top10 <- abund_by_sample_genus %>%
    group_by(Genus) %>%
    summarise(global_abund = sum(total_abund), .groups = "drop") %>%
    arrange(desc(global_abund)) %>%
    slice_head(n = 10) %>%
    pull(Genus)

  df_top <- abund_by_sample_genus %>%
    mutate(Genus = ifelse(Genus %in% top10, Genus, "Other"))

  # Convert to per-sample proportions among included genera (like phage CPM plot)
  prop <- df_top %>%
    group_by(Sample) %>%
    mutate(Proportion = total_abund / sum(total_abund)) %>%
    ungroup()

  # Mean per group
  group_mean <- prop %>%
    group_by(Group, Genus) %>%
    summarise(Mean_Proportion = mean(Proportion), .groups = "drop")

  # Normalize to 100% per group (like phage plot)
  group_mean_norm <- group_mean %>%
    group_by(Group) %>%
    mutate(Mean_Proportion = Mean_Proportion / sum(Mean_Proportion)) %>%
    ungroup()

  if(!include_other){
    group_mean_norm <- group_mean_norm %>%
      filter(Genus != "Other") %>%
      group_by(Group) %>%
      mutate(Mean_Proportion = Mean_Proportion / sum(Mean_Proportion)) %>%
      ungroup()
  }

  # Ordering groups like phage
  group_levels <- c("1 week","1 month","6 months","12 months","mothers","milk")
  group_mean_norm$Group <- factor(group_mean_norm$Group, levels = group_levels)

  # Legend order by total contribution
  genus_order <- group_mean_norm %>%
    group_by(Genus) %>%
    summarise(tot = sum(Mean_Proportion), .groups = "drop") %>%
    arrange(desc(tot)) %>%
    pull(Genus)
  if("Other" %in% genus_order) genus_order <- c("Other", setdiff(genus_order, "Other"))
  group_mean_norm$Genus <- factor(group_mean_norm$Genus, levels = genus_order)

  pal <- ensure_palette(levels(group_mean_norm$Genus), user_palette)
  if("Other" %in% names(pal)) pal["Other"] <- "grey70"

  p <- ggplot(group_mean_norm, aes(Group, Mean_Proportion, fill = Genus)) +
    geom_col(color = "black", linewidth = 0.2) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_manual(values = pal, drop = FALSE) +
    labs(x = NULL, y = "Relative abundance (%)", fill = "Genus", title = title) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(angle = 30, hjust = 1),
      legend.position = "bottom",
      legend.title = element_text(face = "bold")
    )

  list(plot = p, plot_data = group_mean_norm, top10 = top10)
}

plot_all_bacteria <- function(df_bact, meta_df, stool_only = TRUE,
                              title = "Average proportional distribution of bacterial genera (MMC3) — ALL"){
  dat <- base_prepare_bacteria(df_bact, meta_df, stool_only = stool_only)

  abund_by_sample_genus <- dat %>%
    group_by(Sample, Group, Genus) %>%
    summarise(total_abund = sum(Rel_abund_bact), .groups = "drop")

  prop <- abund_by_sample_genus %>%
    group_by(Sample) %>%
    mutate(Proportion = total_abund / sum(total_abund)) %>%
    ungroup()

  group_mean_norm <- prop %>%
    group_by(Group, Genus) %>%
    summarise(Mean_Proportion = mean(Proportion), .groups = "drop") %>%
    group_by(Group) %>%
    mutate(Mean_Proportion = Mean_Proportion / sum(Mean_Proportion)) %>%
    ungroup()

  group_levels <- c("1 week","1 month","6 months","12 months","mothers","milk")
  group_mean_norm$Group <- factor(group_mean_norm$Group, levels = group_levels)

  genus_order <- group_mean_norm %>%
    group_by(Genus) %>%
    summarise(tot = sum(Mean_Proportion), .groups = "drop") %>%
    arrange(desc(tot)) %>%
    pull(Genus)
  group_mean_norm$Genus <- factor(group_mean_norm$Genus, levels = genus_order)

  pal <- ensure_palette(levels(group_mean_norm$Genus), user_palette)

  p <- ggplot(group_mean_norm, aes(Group, Mean_Proportion, fill = Genus)) +
    geom_col(color = "black", linewidth = 0.2) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_manual(values = pal, drop = FALSE) +
    labs(x = NULL, y = "Relative abundance (%)", fill = "Genus", title = title) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(angle = 30, hjust = 1),
      legend.position = "bottom",
      legend.title = element_text(face = "bold")
    )

  list(plot = p, plot_data = group_mean_norm)
}

# ===================== RUN =====================
# Stool-only to be consistent with your ratio plot (recommended)
res_top10_noOther   <- plot_top10_bacteria(bact_genus, meta, stool_only = TRUE, include_other = FALSE,
                                          title = "Average proportional distribution of top-10 bacterial genera (MMC3) — stool only")
res_top10_withOther <- plot_top10_bacteria(bact_genus, meta, stool_only = TRUE, include_other = TRUE,
                                          title = "Average proportional distribution of top-10 bacterial genera (MMC3) — stool only (with Other)")
res_all             <- plot_all_bacteria(bact_genus, meta, stool_only = TRUE,
                                        title = "Average proportional distribution of bacterial genera (MMC3) — stool only (ALL)")

print(res_top10_noOther$plot)

# ===================== SAVE =====================
ggsave(file.path(fig_dir, "bacteria_MMC3_top10_NO_OTHER_stool.png"),
       res_top10_noOther$plot, width = 11, height = 10, dpi = 300, bg = "white")

ggsave(file.path(fig_dir, "bacteria_MMC3_top10_WITH_OTHER_stool.png"),
       res_top10_withOther$plot, width = 11, height = 10, dpi = 300, bg = "white")

ggsave(file.path(fig_dir, "bacteria_MMC3_ALL_stool.png"),
       res_all$plot, width = 11, height = 10, dpi = 300, bg = "white")

# Save plot data (reproducibility)
write_csv(res_top10_noOther$plot_data, file.path(tab_dir, "bacteria_MMC3_top10_NO_OTHER_stool_plotdata.csv"))
write_csv(res_top10_withOther$plot_data, file.path(tab_dir, "bacteria_MMC3_top10_WITH_OTHER_stool_plotdata.csv"))
write_csv(res_all$plot_data, file.path(tab_dir, "bacteria_MMC3_ALL_stool_plotdata.csv"))

# Also save the top10 list used
write_csv(tibble(Genus = res_top10_noOther$top10),
          file.path(tab_dir, "bacteria_MMC3_top10_list_stool.csv"))

```



```{r}
## BASED ON ANNA

suppressPackageStartupMessages({
  library(here)
  library(dplyr); library(readr); library(stringr)
  library(ggplot2); library(scales)
})

# ===================== PATHS =====================
out_dir <- here("results", "figures", "taxonomy_bacteria")
fig_dir <- file.path(out_dir, "figures")
tab_dir <- file.path(out_dir, "tables")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(tab_dir, recursive = TRUE, showWarnings = FALSE)

bact_file <- here("data", "abundance_Genus_B_top10.csv")  # <- TU CSV bacterias
df_b <- read_csv(bact_file, show_col_types = FALSE)

# ===================== PALETTE =====================
user_palette_pastel <- c(
  "Streptococcus"     = "pink",
  "Bacteroides"       = "mediumaquamarine",
  "Escherichia"       = "gold",
  "Faecalibacterium"  = "#C499CC",
  "Bifidobacterium"   = "#6FA3F7",
  "Klebsiella"        = "#7B2CBF",
  "Blautia"           = "#90DDF0",
  "Gemmiger"          = "#FFAC81",
  "Ruminococcus"      = "#FA6459",
  "Pseudomonas"       = "#EEE2DF"
)
user_palette <- user_palette_pastel

# ===================== HELPERS =====================
extract_clean_genus <- function(x){
  g <- ifelse(!is.na(x) & str_detect(x, "g__"),
              str_replace(x, ".*g__", ""),
              x)
  g <- sub("_[A-Z]$", "", g)
  dplyr::recode(g,
    "Kleibsella" = "Klebsiella",
    "Gemminger"  = "Gemmiger",
    .default = g
  )
}

ensure_palette <- function(levels_vec, pal){
  miss <- setdiff(levels_vec, names(pal))
  if(length(miss) > 0){
    pal <- c(pal, setNames(scales::hue_pal()(length(miss)), miss))
  }
  pal
}

# ===================== PREP (BACTERIA) =====================
base_prepare_bacteria <- function(df){
  df %>%
    mutate(
      Group = case_when(
        Origin_time == "infant_7days"   ~ "1 week",
        Origin_time == "infant_1month"  ~ "1 month",
        Origin_time == "infant_6months" ~ "6 months",
        Origin_time == "infant_1year"   ~ "12 months",
        Origin_time == "mother_1month"  ~ "mothers",
        Origin_time == "milk_1month"    ~ "milk",      # <- si existe en tu tabla
        TRUE ~ NA_character_
      ),
      Genus_clean = extract_clean_genus(Genus),
      Abundance = as.numeric(Abundance)
    ) %>%
    filter(!is.na(Group)) %>%
    filter(!is.na(Genus_clean)) %>%
    filter(!is.na(Abundance)) %>%
    filter(Abundance > 0)
}

# ===================== PLOTS =====================
plot_top10_bacteria <- function(df, include_other = FALSE,
                                title = "Average proportional distribution of top-10 bacterial genera"){

  dat <- base_prepare_bacteria(df)

  # Abundance por muestra x genus
  abund_by_sample_genus <- dat %>%
    group_by(Sample, Group, Genus = Genus_clean) %>%
    summarise(total_abund = sum(Abundance), .groups = "drop")

  # top-10 global
  top10 <- abund_by_sample_genus %>%
    group_by(Genus) %>%
    summarise(global_abund = sum(total_abund), .groups = "drop") %>%
    arrange(desc(global_abund)) %>%
    slice_head(n = 10) %>%
    pull(Genus)

  df_top <- abund_by_sample_genus %>%
    mutate(Genus = ifelse(Genus %in% top10, Genus, "Other"))

  # proporción por muestra
  prop <- df_top %>%
    group_by(Sample) %>%
    mutate(Proportion = total_abund / sum(total_abund)) %>%
    ungroup()

  # promedio por grupo (como en fagos)
  group_mean <- prop %>%
    group_by(Group, Genus) %>%
    summarise(Mean_Proportion = mean(Proportion), .groups = "drop")

  # normalizar a 100% por grupo
  group_mean_norm <- group_mean %>%
    group_by(Group) %>%
    mutate(Mean_Proportion = Mean_Proportion / sum(Mean_Proportion)) %>%
    ungroup()

  # quitar Other si quieres y reescalar
  if(!include_other){
    group_mean_norm <- group_mean_norm %>%
      filter(Genus != "Other") %>%
      group_by(Group) %>%
      mutate(Mean_Proportion = Mean_Proportion / sum(Mean_Proportion)) %>%
      ungroup()
  }

  # orden de grupos igual que fagos
  group_levels <- c("1 week","1 month","6 months","12 months","mothers","milk")
  group_mean_norm$Group <- factor(group_mean_norm$Group, levels = group_levels)

  # orden leyenda por contribución total
  genus_order <- group_mean_norm %>%
    group_by(Genus) %>%
    summarise(tot = sum(Mean_Proportion), .groups = "drop") %>%
    arrange(desc(tot)) %>%
    pull(Genus)
  if("Other" %in% genus_order) genus_order <- c("Other", setdiff(genus_order, "Other"))
  group_mean_norm$Genus <- factor(group_mean_norm$Genus, levels = genus_order)

  pal <- ensure_palette(levels(group_mean_norm$Genus), user_palette)
  if("Other" %in% names(pal)) pal["Other"] <- "grey70"

  ggplot(group_mean_norm, aes(Group, Mean_Proportion, fill = Genus)) +
    geom_col(color = "black", linewidth = 0.2) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_fill_manual(values = pal, drop = FALSE, na.translate = FALSE) +
    labs(x = NULL, y = "Relative abundance (%)", fill = "Genus", title = title) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(angle = 30, hjust = 1),
      legend.position = "bottom",
      legend.title = element_text(face = "bold")
    )
}

# ===================== RUN =====================
p_bact_top10_noOther   <- plot_top10_bacteria(df_b, include_other = FALSE)
p_bact_top10_withOther <- plot_top10_bacteria(df_b, include_other = TRUE)

print(p_bact_top10_noOther)

# ===================== SAVE =====================
ggsave(file.path(fig_dir, "bacteria_top10_NO_OTHER.png"),
       p_bact_top10_noOther, width = 11, height = 10, dpi = 300)
ggsave(file.path(fig_dir, "bacteria_top10_WITH_OTHER.png"),
       p_bact_top10_withOther, width = 11, height = 10, dpi = 300)

# (opcional) guardar la tabla usada en el plot
# write_csv(??? , file.path(tab_dir, "bacteria_top10_NO_OTHER_plotdata.csv"))


```


{r}
mycolors<-c( "#FFBB78FF", "#AEC7E8FF","#C5B0D5FF",  "#FF9896FF","#C7C7C7FF" , "#C49C94FF", "#F7B6D2FF", "#98DF8AFF", "#DBDB8DFF", "#9EDAE5FF")
Genus_abundance_top15 <- ggplot(abundance_Genus_B_top10, aes(x = Origin_time, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity", position = "fill", color = NA) +
  labs(x = NULL, y = "Relative Abundance (%)", fill = "Genus") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +  # 🔥 Canvi clau aquí
  scale_fill_manual(values = mycolors) +
  coord_cartesian(ylim = c(0, 1)) +  # Base 1
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5, face = "bold", color = "black"),
    axis.text.y = element_text(face = "bold", color = "black"),
    axis.title.y = element_text(face = "bold", color = "black"),
    legend.position = "right",
    legend.text = element_text(size = 8, color = "black"),
    panel.background = element_rect(fill = "white", color = "grey"),
    plot.background = element_rect(fill = "white", color = "white"),
    panel.grid.major = element_line(color = "grey", size = 0.1),
    panel.grid.minor = element_line(color = "grey", size = 0.1),
    axis.line = element_line(color = "grey", size=0.1),
    axis.ticks = element_line(color = "grey"),
    strip.background = element_rect(fill = "grey", color = "black"),
    strip.text = element_text(face = "bold", color = "black")
  )

Genus_abundance_top15



